Bumble bee competition revisited under global climate change

### 2020-05-25 manuscript_v2.Rmd
added ms v2 on desktop

---
title: Bumble bee competition revisited under global climate change
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
output:
  word_document:
    reference_docx: docx_template.docx
bibliography: refs.bib
---

```{r echo = FALSE, include = FALSE}
### Prepare environment
library(bipartite)
library(lme4)
library(lmerTest)
library(tidyverse)
library(lubridate)

### Load site data
site_data <- read_csv("./processed_data/site_data.csv") %>%
  dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean)) %>% # drop these variables
  mutate(elev.class = factor(elev.class, levels = c("oben", "mitte", "unten")),
         elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) # turn elav.class into an ordered factor

elev_classification <- site_data %>% 
  select(elev.mean, elev.class) %>% 
  unique() %>% 
  arrange(-elev.mean)

### Climate data
gdd <- read_csv("./processed_data/climate.csv") %>%
  mutate(year = factor(year)) %>%
  select(site, date, gdd.cum)

### Load network and survey data
bb_traits <- read_csv("./processed_data/bb_traits.csv") %>%
  select(subgenus, bb.sp, pbl.w, pbl.w.class)

net <- read_csv("./processed_data/network.csv") %>%
  unite(site.year, site, year, sep = "_", remove = FALSE) %>% # create site.year field that can later be split back up
  unite(site.date, site, date, sep = "_", remove = FALSE) %>%
  mutate(year = factor(year)) %>% # convert year to factor
  full_join(bb_traits, by = c("bb.sp.lat" = "bb.sp")) %>%
  left_join(site_data, by = "site") %>%
  left_join(gdd, by = c("site", "date")) %>%
  arrange(site, date)

survey <- read_csv("./processed_data/floral_survey.csv") %>%
  mutate(year = factor(year)) %>% # convert year to factor
  semi_join(net, by = c("site", "plant.sp")) %>% # Consider only plants visited at a given site-date
  left_join(gdd, by = c("site", "date")) %>%
  arrange(site, date)

### Per-transect Bombus density
bb_dens <- net %>%
  group_by(site, date, elev.class, elev.mean, year) %>%
  summarize(bb.dens = n())

### Per-transect floral density
fl_dens <- survey %>%
  group_by(site, date, year) %>%
  summarize(fl.dens = sum(flower.cover),
            log.fl.dens = log(sum(flower.cover)))

### Join BB density and floral cover
density <- bb_dens %>%
  full_join(fl_dens) %>% # Note that there is not perfect alignment in site-dates between the network observations and the floral surveying, so this join will cause some NAs that will need to be removed, causing some transects to drop out of the dataset
  na.omit() %>%
  ungroup %>%
  mutate(bb_per_fl = bb.dens/fl.dens,
         site = factor(site),
         yday = yday(date),
         week = week(date),
         month = month(date))

density_month.agg <- bb_dens %>%
  full_join(fl_dens) %>% # Note that there is not perfect alignment in site-dates between the network observations and the floral surveying, so this join will cause some NAs that will need to be removed, causing some transects to drop out of the dataset
  na.omit() %>%
  ungroup %>%
  mutate(site = factor(site),
         yday = yday(date),
         week = week(date),
         month = month(date)) %>%
  group_by(year, month, site, elev.mean, elev.class) %>%
  summarize(mean.bb.dens = mean(bb.dens),
            mean.fl.dens = mean(fl.dens)) %>%
  mutate(log.mean.fl.dens = log(mean.fl.dens),
         bb.per.fl = mean.bb.dens/mean.fl.dens)

###############################
######### Define functions
##############################

# Find breakpoints on GDD scale
breakpoints_gdd <- function(x, year, t1, t2) {

  data <- filter(x, year == !!year)

  # Logistic regression for first break point (founding to buildup)
  glm1.low <- glm(q.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm1.mid <- glm(q.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm1.high <- glm(q.prop ~ gdd.cum,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  # Logistic regression for second break point (buildup to reproductive)
  glm2.low <- glm(r.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm2.mid <- glm(r.prop ~ gdd.cum,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm2.high <- glm(r.prop ~ gdd.cum,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  ### Extract gdd corresponding to desired proportion; based on https://stackoverflow.com/questions/32040504/regression-logistic-in-r-finding-x-value-predictor-for-a-particular-y-value

  getX <- function(x, y) {
    (log(y/(1-y)) - coef(x)[1])/coef(x)[2]
    }

  break1.low <- getX(glm1.low, t1) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break1")

  break1.mid <- getX(glm1.mid, t1) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break1")

  break1.high <- getX(glm1.high, t1) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break1")

  break2.low <- getX(glm2.low, t2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break2")

  break2.mid <- getX(glm2.mid, t2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break2")

  break2.high <- getX(glm2.high, t2) %>%
    data.frame() %>%
    select(gdd = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break2")

  # Join into data frame, convert elev.class2 to ordered factor, spread
  breaks <- full_join(break1.low, break1.mid) %>%
    full_join(break1.high) %>%
    full_join(break2.low) %>%
    full_join(break2.mid) %>%
    full_join(break2.high) %>%
    mutate(elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) %>%
    spread(breakID, gdd)

  #
  out <- data %>%
    full_join(breaks) %>%
    gather(caste, count, -c(site, date, year, q.prop, r.prop,
                          gdd.cum, yday, elev.class, elev.class2, 
                          management, transect, elev.mean, 
                          lat, lon, break1, break2)) %>%
  group_by(site, date) %>%
  mutate(prop = count/sum(count))
}

# Find breakpoints on Julian scale
breakpoints_yday <- function(x, year, t1, t2) {

  data <- filter(x, year == !!year)

  # Logistic regression for first break point (founding to buildup)
  glm1.low <- glm(q.prop ~ yday,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm1.mid <- glm(q.prop ~ yday,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm1.high <- glm(q.prop ~ yday,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  # Logistic regression for second break point (buildup to reproductive)
  glm2.low <- glm(r.prop ~ yday,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "low"))

  glm2.mid <- glm(r.prop ~ yday,
                  family = quasibinomial,
                  data = filter(data, elev.class2 == "mid"))

  glm2.high <- glm(r.prop ~ yday,
                   family = quasibinomial,
                   data = filter(data, elev.class2 == "high"))

  ### Extract gdd corresponding to desired proportion; based on https://stackoverflow.com/questions/32040504/regression-logistic-in-r-finding-x-value-predictor-for-a-particular-y-value

  getX <- function(x, y) {
    (log(y/(1-y)) - coef(x)[1])/coef(x)[2]
    }

  break1.low <- getX(glm1.low, t1) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break1")

  break1.mid <- getX(glm1.mid, t1) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break1")

  break1.high <- getX(glm1.high, t1) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break1")

  break2.low <- getX(glm2.low, t2) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break2")

  break2.mid <- getX(glm2.mid, t2) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "mid",
           breakID = "break2")

  break2.high <- getX(glm2.high, t2) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "high",
           breakID = "break2")

  # Join into data frame, convert elev.class2 to ordered factor, spread
  breaks <- full_join(break1.low, break1.mid) %>%
    full_join(break1.high) %>%
    full_join(break2.low) %>%
    full_join(break2.mid) %>%
    full_join(break2.high) %>%
    mutate(elev.class2 = factor(elev.class2, levels = c("high", "mid", "low"))) %>%
    spread(breakID, day)

  #
  out <- data %>%
    full_join(breaks) %>%
    gather(caste, count, -c(site, date, year, q.prop, r.prop,
                          gdd.cum, yday, elev.class, elev.class2, 
                          management, transect, elev.mean, 
                          lat, lon, break1, break2)) %>%
  group_by(site, date) %>%
  mutate(prop = count/sum(count))
}

# Find Juian breakpoints across all elevations
breakpoints_yday_all <- function(x, year, t1, t2) {

  data <- filter(x, year == !!year)

  # Logistic regression for first break point (founding to buildup)
  glm1 <- glm(q.prop ~ yday,
                  family = quasibinomial,
                  data = data)

  # Logistic regression for second break point (buildup to reproductive)
  glm2 <- glm(r.prop ~ yday,
                  family = quasibinomial,
                  data = data)

  ### Extract day corresponding to desired proportion; based on https://stackoverflow.com/questions/32040504/regression-logistic-in-r-finding-x-value-predictor-for-a-particular-y-value

  getX <- function(x, y) {
    (log(y/(1-y)) - coef(x)[1])/coef(x)[2]
    }

  break1 <- getX(glm1, t1) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break1")

  break2 <- getX(glm2, t2) %>%
    data.frame() %>%
    select(day = 1) %>%
    mutate(elev.class2 = "low",
           breakID = "break2")

  # Join into data frame, convert elev.class2 to ordered factor, spread
  breaks <- full_join(break1, break2) %>%
    spread(breakID, day)

  #
  out <- data %>%
    full_join(breaks) %>%
    gather(caste, count, -c(site, date, year, q.prop, r.prop,
                          gdd.cum, yday, elev.class, elev.class2, 
                          management, transect, elev.mean, 
                          lat, lon, break1, break2)) %>%
  group_by(site, date) %>%
  mutate(prop = count/sum(count))
}

# Binomial ggplot smoother
binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = FALSE, ...)
}
```

@@

The nascent reality of climate change has already revealed the fragile contingency of ecological paradigms that rely on the implicit assumption of climatic stability [@Hannah2002-nv; @Walther2002-mq]. Thus, one of the central questions facing ecologists today is how climate change will affect the future fitness and distribution of species and, thereby, the composition and function of ecological communities. While it is intuitive that climate change should affect species and communities through environmental filtering, it is now recognized that climate change can also restructure biotic interactions [@Tylianakis2008-fi], such as competition and facilitation [@Ettinger2017-pk], altering the fitness landscape that governs community assembly and species coexistence [@Chesson2000-aw].

Bumble bees (Hymenoptera: *Bombus* spp.) have become a model system for the study of biotic interactions in ecological communities. In particular, the broad and overlapping distributions of bumble bee species, together with their gross similarity in morphology, behavior, and dietary niche, have attracted the attention of researchers interested in questions of interspecific competition and, more broadly, the rules of community assembly and species coexistence [@Plowright1984-lc; @Goulson2008-zy]. In this regard, however, bumble bees have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble populations [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communties have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phyologeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [e.g. @Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous "mosaic" of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread "core species", and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-xd].

As pointed out by @Williams1989-vr, however, the question of competition as a biogeographic/evolutionary-scale driver of communty assembly differs from that of competition as a local/behavioral-scale driver of species' interactions. While the role of competition in bumble bee community assembly remains somewhat obscure, there is broad empirical evidence that bumble bees facultatively partition resources on the basis of trait-matching and possibly dominance hierarchy [e.g @Inouye1978-qi; @Morse1982-bh; @Ishii2013-sm; @Brosi2013-mu]. This distinction between spatiotemporal scales of competition, echoing the logic of @Ranta1981-vx, becomes especially important when the scope of inquiry switches from retrospective to prospective, and all the more so when there is reason to think that the ecological context of competition is undergoing rapid and unprecedented change. Since the height of bumble bee competition research in the 1980s, such a shift has indeed occurred, with retrospective questions of community assembly eclipsed by the urgency of conservation in the face of global changes in climate and land use [@Goulson2010-lo]. Given the uncertain spatiotemporal lags and exogenous drivers of community assembly, inferring the past role of competition through the study of bumble bee community structure may be of limited use in predicting the future fitness and distribution of bumble bee species, together with the ecological functions and services they provide. Studying local/behavioral-scale competition in bumble bee communities along climate gradients provides a vital complement to the classical study of bumble bee community structure.

The study of competition at local/behavioral scales, however, entails two empirical challenges. First, it is not enough simply to know which species occur and their similarity in traits/phyologeny; one must record their behavior with respect to their shared resources, and at a frequency appropriate to the temporal scale of inquiry. Second, climate variation must be decoupled from confounding spatial effects. The use of species interaction networks along steep elevation gradients provides an elegant solution to both of these challenges.

Interaction networks provide a tractable analytical framework for the study of competition and resource partitioning [], and steep elevation gradients can serve as space-for-time proxies in the study of climate effects, capturing a broad range of climate variation while minimizing the confounding effects of geographic distance [e.g. @Hoiss2015-gv].

Our study is set in the Berchtesgadener Alps of southwestern Germany, where a steep elevation gradient permits the study of wide environmental variability while minimizing the confounding effects of geographical distance [e.g. @Hoiss2015-gv]. In this context, we formulate a "metahypothesis" concerning the variation of hypothesized competition processes through elevation and time (**Figure 1**). Consistent with the stress-gradient hypothesis [@Bertness1994-zx] and with previous empirical work in our study system [@Hoiss2012-sn], we hypothesize an inverse relationship between competition and elevation. Along a temporal axis, we hypothesize that the strength of competition is structured by the seasonal life history pattern of bumble bee colonies [@Goulson2010-lo]. During the solitary phase, when the bumble bee community consists only of foundress queens, competition should be low. Both inter- and intraspecific competition should intensify as colonies increase in size during the social phase. After the switch point, when colonies begin producing gynes and males instead of workers, intraspecific comeptition should decrease due to the strong morphological differences between queens, males, and workers within each species, but interspecific competition might intensify due to the effective "over-packing" of trait space wih the introduction of multiple morphological castes per species.

In addition to clarifying the basic ecology of bumble bee community assembly, parsing competitive processes in this way has an applied significance for the conservation of bumble bees in the changing environments of the Anthropocene. Under the forcing of environmental change, such as climate warming or land use conversion, bumble bee communities might not be affected only in terms of range shifts, but also by a restructuring of competitive interactions, the nature and consequences of which may be difficult to predict.

One of the most well-studied hypotheses linking climate and biotic interactions is the "stress gradient hypothesis", which posits a that the relative significance of competition and faciitation vary predictably along gradients of abiotic (e.g. climatic) stress, with low stress environments characterized by competition and high stress environments by facilitation [@Bertness1994-zx].

To account for both the importance and variability of competitive processes in bumble bee communities, it is helpful to distinguish three processes that could obtain to varying degrees: no competition, interspecific competition, and intraspecific competition (**Table 1**). We will refer to these processes as hypotheses, since their realization in any given community is matter of inference. Below, we elaborate each and formulate corresponding empirical predictons.

+-------------------------------+----------------------------------------------------+
| Hypothesis                    | Prediction                                         |
+===============================+====================================================+
| H1: No competition            | 1. BB density !~ FL density                        |
|                               | 2. BB trait dispersion = null expectation          |
|                               | 3. BB-FL trait-matching > null expectation         |
+-------------------------------+----------------------------------------------------+
| H2: Interspecific competition | 1. BB density ~ FL density                         |
|                               | 2. BB trait dispersion > null expectation          |
|                               | 3. BB-FL trait-matching > null expectation         |
+-------------------------------+----------------------------------------------------+
| H3: Intraspecific competition | 1. BB density ~ FL density                         |
|                               | 2. BB diet breadth ~ BB conspecific-packing        |
|                               | 3. BB-FL trait-matching ~ 1/BB conspecific-packing |
+-------------------------------+----------------------------------------------------+

@@

*H1: No competition.*    The hypothesis that floral resource competition is not a significant process governing a bumble bee community functions as a "null hypothesis" when competition *per se* is the phenomenon in question, but it is not "null" in the sense of mere negation of mechanistic explanation. To the contrary, at least two plausible and ecologically meaningful mechanisms can be proposed that would account for a lack of competition in bumble bee communities [@Ranta1981-vx]. First, it is possible that a super-abundance of floral resources could saturate the foraging capacity of a local bumble bee population. Indeed, this may be a common (though transient) phenomenon associated with the bloom of mass flowering crops, such as red clover [@Ranta1981-vx] and oilseed rape [@Westphal2009-sb]. Alternatively, competition could be attenuated if abiotic stress, such as low temperature, suppresses bumble bee populations below competition-inducing densities. This pattern would be a prediction of the "stress-gradient hypothesis" [@Bertness1994-zx], and while it has not, to our knowledge, been documented for bumble bees specifically, a recent study of whole bee communities (of which bumble bees were a part) found evidence of dampened competition at high elevation sites along a montane-to-alpine gradient [@Hoiss2012-sn].

Under either mechanism, the most straightforward prediction is that bumble bee population density would be decoupled from resource density, since the latter would be non-limiting. For the same reason, the absence of competition would also predict that the trait dispersion of sympatric bumble bee species would not exceed that predicted by a random null model. Third, under a no-competition scenario, bumble bee species would suffer no displacement from their optimal floral resource species. Thus, holding other aspects of floral resource value constant (e.g. abundance, nectar volume/concentration), a lack of competition would predict that morphological trait-matching between bumble bee species and their chosen host flowers would exceed that predicted by a random null model.

*H2: Interspecific competition.*    Resource partitioning by bumble bees is mediated by variation in matching traits, the most well-studied of these being tongue length. Bumble bee tongue lengths vary between species, and the efficiency with which a bumble bee can collect nectar from a given flower depends largely on the closeness of match between the tongue length of the bee and the corolla depth of the flower [@Inouye1980-mv; @Ranta1980-ok]. Thus, under conditions of interspecific competition, a species' "defensible" niche space (i.e. the space in which it is a superior competitor) should center on the floral trait space that corresponds to its tongue length, and the number of species that can be "packed" into a given locality should be limited by some minimal differentiation of tongue length [@Inouye1977-kl; @Pyke1982-gn] within the range of variation supported by the trait space of available flora.

Since resource limitation is a necessary condition for resource competition, a key prediction distiguishing interspecific competition from a lack of competition is that bumble bee density should vary in proportion to floral resource density. A bumble bee community under interspecific competition should also exhibit higher tongue length dispersion and stronger trait matching relative to the predictions of a random null model [@Inouye1977-kl].

*H3: Intraspecific competition.*    As noted by @Ranta1981-vx, theory predicts that intraspecific competition can promote species coexistence through the relief of interspecific competition [@MacArthur1984-qz], and could thus, as an alternative to the "no competition" hypothesis, account for cases in which the observed diversity of sympatric bumble bee species exceeds that predicted by trait-based niche models. Under the pressure of intraspecific competition, individuals face the depletion of preferred resources and may respond by foraging more opportunistically within the constraints of their fundamental niche space []. Alternatively, a population may relieve intraspecific competition by subdividing, on the basis of intraspecific trait heterogeneity, into differently specialized "guilds" [@Araujo2008-bp]. In either case, the population-level effect would be an increase in the species' realized niche breadth [though see @Parent2014-hx].

Thus, the predictions of intraspecific competition can be formulated as follows. First, as in interspecific competition, resources must be limiting, so bumble bee density should vary in proportion to floral density. In the case of intraspecific competition, though, we can refine this prediction to say that the density of a given bumble bee species should vary in proportion to the density of the portion of the floral resource community belonging to its dietary niche, since the effects of intraspecific competition are realized on a per-species basis. Second, the dietary breadth of a given species should vary in proportion to the density of conspecifics per density of floral resources (we will refer to this as a "conspecific-packing" index). Conversely, a broadening of diet should entail a relaxation of trait-matching, so the strength of trait-matching should vary inversely with conspecific-packing.

@@

```{r echo = FALSE, figu}
Time <- c(0,1,2,3)
Elevation <- c(0,1,2,3)
tab <- cbind(Time, Elevation) %>%
  as_tibble()

ggplot(tab, aes(Time, Elevation)) +
  geom_vline(xintercept = c(1,2), color = "white") +
  geom_hline(yintercept = c(1,2), color = "white") +
  scale_x_continuous(expand = c(0,0),
                     breaks = c(0.5, 1.5, 2.5), 
                     labels = c("founding", "build-up", "reproductive"),
                     limits = c(0,3)) +
  scale_y_continuous(expand = c(0,0),
                     breaks = c(0.5, 1.5, 2.5), 
                     labels = c("low", "mid", "high"),
                     limits = c(0,3)) +
  #geom_text() +
  annotate("text", x = 0.5, y = 2.5, label = "H1") +
  annotate("text", x = 1.5, y = 2.5, label = "H1") +
  annotate("text", x = 2.5, y = 2.5, label = "H1") +
  annotate("text", x = 0.5, y = 1.5, label = "H1") +
  annotate("text", x = 0.5, y = 0.5, label = "H1") +
  annotate("text", x = 1.5, y = 1.5, label = "H2 | H3") +
  annotate("text", x = 2.5, y = 1.5, label = "H2 < H3") +
  annotate("text", x = 1.5, y = 0.5, label = "H2 | H3") +
  annotate("text", x = 2.5, y = 0.5, label = "H2 < H3") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

**Figure 1:** Hypothesized variation in prevailing competitive processes through time and elevation. \

@@

*Field sampling*

@@

*Inferring reosurce limitation*

**Results**\

>>>>

---
title: "Bumble bee competition revisited under global climate change"
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter
output:
  word_document:
    reference_docx: docx_template.docx
  pdf_document: default
bibliography: refs.bib
---

```{r echo = FALSE, include = FALSE}
library(tidyverse)
```

@@

The nascent reality of climate change has already revealed the fragile contingency of ecological paradigms that rely on the implicit assumption of climate stability [@Hannah2002-nv; @Walther2002-mq]. Thus, one of the central questions facing ecologists today is how climate change will affect the future fitness and distribution of species and, thereby, the composition and function of ecological communities. While it is intuitive that climate change should affect species and communities through environmental filtering, it is now recognized that climate change can also restructure biotic interactions [@Tylianakis2008-fi], such as competition and facilitation [@Ettinger2017-pk], altering the fitness landscape that governs community assembly and species coexistence [@Chesson2000-aw].

Bumble bees (Hymenoptera: *Bombus* spp.) have become a model system for the study of biotic interactions in ecological communities. In particular, the broad and overlapping distributions of bumble bee species, together with their gross similarity in morphology, behavior, and dietary niche, have attracted the attention of researchers interested in questions of interspecific competition and, more broadly, the rules of community assembly and species coexistence [@Plowright1984-lc; @Goulson2008-zy]. In this regard, however, bumble bees have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble populations [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communties have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phyologeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [e.g. @Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread "core species", and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-xd].

As pointed out by @Williams1989-vr, however, the question of competition as a biogeographic/evolutionary-scale driver of communty assembly differs from that of competition as a local/behavioral-scale driver of species' interactions. While the role of competition in bumble bee community assembly remains somewhat obscure, there is broad empirical evidence that bumble bees facultatively partition resources to minimize competition [e.g @Inouye1978-qi; @Morse1982-bh; @Ishii2013-sm; @Brosi2013-mu; @Brosi2017-oc]. This distinction between spatiotemporal scales of competition, echoing the logic of @Ranta1981-vx, becomes especially important when the scope of inquiry switches from retrospective to prospective, and all the more so when there is reason to think that the ecological context of competition is undergoing rapid and unprecedented change. Since the height of bumble bee competition research in the 1980s, such a shift has indeed occurred, with retrospective questions of community assembly eclipsed by the urgency of conservation in the face of global changes in climate and land use [@Goulson2010-lo]. Given the uncertain spatiotemporal lags and exogenous drivers of community assembly, inferring the past role of competition through the study of bumble bee community structure may be of limited use in predicting the future fitness and distribution of bumble bee species. In this regard, the study of competition via behavioral patterns of floral resource partitioning that respond in real-time to variation in ecological context [@Spiesman2016-nf] provides a vital complement to the classical study of community assembly.

The study of competition as a behavioral phenomenon, however, is empirically and analytically demanding, since species interactions, not merely community composition, must be measured and interpreted. The difficulty is compounded when the goal is to infer the effects of environmental variation, since observations must be replicated across gradients of interest while minimizing the influence of confounding spatial or temporal effects. An elegant solution to both these challenges is the use of species interaction networks along steep elevation gradients [e.g. @Hoiss2015-gv; @Albrecht2018-gn]. Interaction network analysis provides a tractable framework for the study of resource partitioning [@Bluthgen2006-if], and steep elevation gradients can serve as space-for-time proxies in the study of climate effects, capturing a broad range of climate variation while minimizing the confounding effects of geographic distance [@Blois2013-wr]. To our knowledge, however, the analysis of plant-pollinator interaction networks along elevational climate gradients has never been applied specifically to the question of competition in bumble bees and its relation to climate change.

In this study, we analyze patterns of resource partitioning among bumble bees along an elevation gradient in the Berchtesgadener Alps of southeastern Germany. Our analysis has a hierarchical structure. First, we test the hypothesis that competition is of negligible significance in our study system. At least two plausible mechanisms can be proposed that would account for a lack of competition in bumble bee communities [@Ranta1981-vx]. First, it is possible that a super-abundance of floral resources could saturate the foraging capacity of a local bumble bee population. Indeed, this may be a common (though transient) phenomenon associated with the bloom of mass flowering crops, such as red clover [@Ranta1981-vx] and oilseed rape [@Westphal2009-sb]. Alternatively, competition could be attenuated if abiotic stress, such as low temperature, suppresses bumble bee populations below competition-inducing densities. This pattern would be a prediction of the "stress-gradient hypothesis" [@Bertness1994-zx], and while it has not, to our knowledge, been documented for bumble bees specifically, a recent study of whole bee communities in the Berchtesgaden system (of which bumble bees were a part) found evidence of dampened competition at high elevation sites along a montane-to-alpine gradient [@Hoiss2012-sn]. Under either mechanism, the most straightforward prediction is that bumble bee population density would be decoupled from resource density, since the latter would be non-limiting. For the same reason, the absence of competition would also predict that resource partitioning would be constrained only by trait-matching, not by niche overlap; thus, niche partitioning among species with similar traits should not exceed that predicted by a random null model.

Where the hypothesis of no competition can be rejected, we seek to differentiate inter- from intraspecific competition. These two forms of competition share the necessary precondition of resource limitation, but they imply distinct predictions regarding resource partitioning and trait-matching. Resource partitioning by bumble bees is mediated by variation in matching traits, the most well-studied of these being tongue length. Bumble bee tongue lengths vary between species, and the efficiency with which a bumble bee can collect nectar from a given flower depends largely on the closeness of match between the tongue length of the bee and the corolla depth of the flower [@Inouye1980-mv; @Ranta1980-ok]. Since tongue length also varies within species, and literature values for a given species vary, it is common to average available values and bin species into tongue-length "guilds" [@Pellissier2013-pk].

Where interspecific competition is strong, a species' dietary specialization should increase in proportion to the density of other species within the same tongue length guild per density of floral resources shared by that guild (hereafter referred to as "guild-packing"). Moreover, species within a tongue length guild should tend to exhibit divergent abundance dynamics, since they are reciprocally limiting. Under the pressure of intraspecific competition, however, individuals may respond by foraging more opportunistically within the constraints of their fundamental niche space []. Alternatively, a species may relieve intraspecific competition by subdividing, on the basis of intraspecific trait heterogeneity, into differently specialized populations [@Araujo2008-bp]. In either case, the species-level effect would be an increase in realized niche breadth (i.e. a decrease in specialization) in proportion to conspecific-packing (density of conspecifics per density of guild-level floral resources) [though see @Parent2014-hx].

These hypotheses and predictions are summarized in Table 1.

+--------------------------------+--------------------------------------------------------+
| Hypothesis                     | Prediction                                             |
+================================+========================================================+
| H1: No competition             | 1. BB density !~ FL density                            |
+--------------------------------+--------------------------------------------------------+
| H2a: Interspecific competition | 1. specialization ~ guild-packing                      |
|                                | 2. within-guild abundance divergence                   | 
+--------------------------------+--------------------------------------------------------+
| H2b: Intraspecific competition | 1. specialization ~ 1/conspecific-packing              |
|                                | 2. trait-matching ~ 1/conspecific-packing              |
+--------------------------------+--------------------------------------------------------+

@@

In this context, we formulate a "metahypothesis" concerning the variation of hypothesized competition processes through elevation and time (**Figure 1**). Consistent with the stress-gradient hypothesis [@Bertness1994-zx] and with previous empirical work in our study system [@Hoiss2012-sn], we hypothesize an inverse relationship between competition and elevation. Along a temporal axis, we hypothesize that the strength of competition is structured by the seasonal life history pattern of bumble bee colonies [@Goulson2010-lo]. During the solitary phase, when the bumble bee community consists only of foundress queens, competition should be low. Both inter- and intraspecific competition should intensify as colonies increase in size during the social phase. After the switch point, when colonies begin producing gynes and males instead of workers, intraspecific comeptition should decrease due to the strong morphological differences between queens, males, and workers within each species, but interspecific competition might intensify due to the effective "over-packing" of trait space wih the introduction of multiple morphological castes per species.

@@

```{r echo = FALSE, fig.cap = "**Figure 1:** Hypothesized variation in prevailing competitive processes through time and elevation."}
Time <- c(0,1,2,3)
Elevation <- c(0,1,2,3)
tab <- cbind(Time, Elevation) %>%
  as_tibble()

ggplot(tab, aes(Time, Elevation)) +
  geom_vline(xintercept = c(1,2), color = "white") +
  geom_hline(yintercept = c(1,2), color = "white") +
  scale_x_continuous(expand = c(0,0),
                     breaks = c(0.5, 1.5, 2.5), 
                     labels = c("founding", "build-up", "reproductive"),
                     limits = c(0,3)) +
  scale_y_continuous(expand = c(0,0),
                     breaks = c(0.5, 1.5, 2.5), 
                     labels = c("low", "mid", "high"),
                     limits = c(0,3)) +
  #geom_text() +
  annotate("text", x = 0.5, y = 2.5, label = "H1") +
  annotate("text", x = 1.5, y = 2.5, label = "H1") +
  annotate("text", x = 2.5, y = 2.5, label = "H1") +
  annotate("text", x = 0.5, y = 1.5, label = "H1") +
  annotate("text", x = 0.5, y = 0.5, label = "H1") +
  annotate("text", x = 1.5, y = 1.5, label = "H2 | H3") +
  annotate("text", x = 2.5, y = 1.5, label = "H2 < H3") +
  annotate("text", x = 1.5, y = 0.5, label = "H2 | H3") +
  annotate("text", x = 2.5, y = 0.5, label = "H2 < H3") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```
\

@@

*Field sampling.*  Our study is set in the Berchtesgadener Alps of southwestern Germany, a system in which both plant and pollinator communities have been well-studied [@Hoiss2012-sn; Hoiss2012-kr; Hoiss2015-gv]. Our study sites are the same as those  ranging in elevation from 641 to 2032 m above sea level.

@@

*Colony life cycle analysis.* Bumble bee colonies exhibit three distinct life history phases: a "founding" phase during which colonies are initiated by solitary queens, a "build-up" phase during which workers are produced, and a "reproductive" phase where colonies produce new queens and males [@Goulson2010-lo]. A formal delimitation of these phases serves both as a temporal framework for analyzing bumble bee and floral dynamics and as a check that different elevational subsets of our data capture the same life history interval and are thus comparable in terms of network structure.

To delimit colony phases, we fit binomial regression models to the proportional abundance of foundress queens (Q) and new reproductives (R) (males + new queens), respectively. We set the breakpoint between the founding and build-up stages at Q = 0.75, and the breakpoint between buildup and reproductive stages at R = 0.1. These cutoffs are somewhat arbitrary points in a continuous process, but their primary purpose is to facilitate comparison across sites. This analysis was performed on both a GDD and Julian temporal scale. All operations were done in R.
